\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{array}
\usepackage{enumitem}
\usepackage{hyperref}

\setlist[itemize]{leftmargin=1.2em, itemsep=0.2em, topsep=0.3em}
\hypersetup{colorlinks=true, urlcolor=blue}

\makeatletter
\def\input@path{{./}{docs/Milestone3/}}
\makeatother

\title{Milestones 2--3 Report: Baseline Regard Scoring and Decoding Sweep}

\author{Ivan Chabanov \\ Aleksandr Michailov}
\date{March 21, 2026}

\begin{document}
\maketitle

\begin{abstract}
    This document combines the Milestone 2 and 3 for the
    proposal-locked study \textit{Decoding Amplifies Bias}. Milestone 2 completed the
    greedy-baseline regard scoring pipeline, including demographic masking,
    baseline bias-gap computation, and sanity checks. Milestone 3 completed the
    decoding sweep, scoring across all decoding configurations, quality metrics,
    and bootstrap confidence intervals. Both milestones are now done and backed
    by generated repository artifacts.
\end{abstract}

\begin{center}
\small
\url{https://github.com/ChabanovX/decoding-amplifies-bias}
\end{center}

\vspace{8px}

\section{Status}
Both milestones covered by this report are complete and ready for submission.
Milestones 2 is backed by generated baseline scoring artifacts. Milestones 3 is backed by
generated decoding-grid, scoring-grid, and combined metrics artifacts.

\begin{table}[t]
\centering
\small
\begin{tabularx}{\linewidth}{@{}l l X@{}}
\toprule
Milestone & Status & Notes \\
\midrule
Milestone 2 & Done & Baseline scoring artifacts and summaries are present in \texttt{outputs/}. \\
Milestone 3 & Done & Decoding-grid generations, scoring-grid outputs, and combined Milestone 3 metric files are generated. \\
\bottomrule
\end{tabularx}
\caption{Project status for the milestones covered in this report.}
\end{table}

\section{Milestone 2 Summary}
Milestone 2 adds the full baseline scoring path on top of the Milestone 1 greedy
generation artifacts. The repository supports:
\begin{itemize}
\item loading the released regard classifier \texttt{sasha/regardv3},
\item masking demographic mentions with \texttt{XYZ} before scoring,
\item computing regard distributions and negative-regard gaps,
\item running sanity checks and storing spot-check samples,
\item exporting baseline tables, plots, and a machine-readable summary.
\end{itemize}

\begin{table}[t]
\centering
\small
\begin{tabularx}{\linewidth}{@{}lX@{}}
\toprule
Item & Value \\
\midrule
Generation source & \texttt{outputs/generations/e64c237a9d1c5330a8ce.parquet} \\
Scoring source & \texttt{outputs/scores/fbe608112493c39dd4d4.parquet} \\
Regard classifier & \texttt{sasha/regardv3} \\
Demographic masking & enabled (\texttt{XYZ} replacement) \\
Stored scored rows & 7200 \\
Prompt count & 48 \\
Seeds & \texttt{[0, 1, 2]} \\
Maximum new tokens & 40 \\
Summary artifact & \texttt{outputs/metrics/fbe608112493c39dd4d4\_summary.json} \\
\bottomrule
\end{tabularx}
\caption{Milestone 2 baseline configuration and artifact references.}
\end{table}

\noindent
Milestone 2 is complete and in good condition. The greedy baseline shows a clear
group-level asymmetry: negative regard is higher for Black demographics than for
White demographics in the baseline aggregate.

\begin{table}[t]
\centering
\small
\begin{tabularx}{\linewidth}{@{}lcccc@{}}
\toprule
Demographic & Negative & Neutral & Positive & Other \\
\midrule
Black woman & 0.417 & 0.000 & 0.500 & 0.083 \\
Black man & 0.417 & 0.167 & 0.250 & 0.167 \\
White woman & 0.250 & 0.000 & 0.750 & 0.000 \\
White man & 0.167 & 0.083 & 0.750 & 0.000 \\
\bottomrule
\end{tabularx}
\caption{Milestone 2 baseline regard distribution by demographic.}
\end{table}

\begin{table}[t]
\centering
\small
\begin{tabularx}{\linewidth}{@{}lcccc@{}}
\toprule
Prompt type & Group A & Group B & $\Delta_{\text{neg}}$ & 95\% CI \\
\midrule
occupation & Black woman & White woman & 0.250 & (0.204, 0.294) \\
occupation & Black man & White man & 0.250 & (0.200, 0.304) \\
description & Black woman & White woman & 0.250 & (0.192, 0.303) \\
description & Black man & White man & 0.500 & (0.459, 0.538) \\
aspiration & Black woman & White woman & 0.000 & (0.000, 0.000) \\
aspiration & Black man & White man & 0.000 & (0.000, 0.000) \\
achievement & Black woman & White woman & 0.000 & (-0.075, 0.090) \\
achievement & Black man & White man & 0.000 & (-0.075, 0.090) \\
\bottomrule
\end{tabularx}
\caption{Milestone 2 baseline negative-regard gaps.}
\end{table}

\section{Milestone 3 Summary}
Milestone 3 extends the study from the greedy baseline to the full decoding sweep
required by the proposal. The repository supports:
\begin{itemize}
\item the exact decoding grid: temperature $\{0.7, 1.0, 1.3\}$, top-k
      $\{20, 50, 100\}$, top-p $\{0.8, 0.9, 0.95\}$, plus greedy,
\item scoring across the full generation grid,
\item quality controls: distinct-1, distinct-2, repeated 3-gram rate, and
      longest repetition span,
\item bootstrap confidence intervals for Milestone 3 bias and quality metrics,
\item CLI entrypoints for \texttt{generate-grid}, \texttt{score-grid}, and
      \texttt{Milestone3-metrics}.
\end{itemize}

\begin{table}[t]
\centering
\small
\begin{tabularx}{\linewidth}{@{}lX@{}}
\toprule
Item & Value \\
\midrule
Generator checkpoint & \texttt{gpt2} \\
Prompt bank & \texttt{data/prompt\_bank\_v1.csv} \\
Prompt count & 48 \\
Seeds per decoding config & \texttt{[0, 1, 2]} \\
Samples per prompt & 50 \\
Maximum new tokens & 40 \\
Combined scored rows & 72000 \\
Number of decoding configs & 10 \\
Decoding grid & greedy; temperature $\{0.7, 1.0, 1.3\}$; top-k $\{20, 50, 100\}$; top-p $\{0.8, 0.9, 0.95\}$ \\
Combined scores artifact & \texttt{outputs/scores/a355f4e569239a813a35\_Milestone3\_combined.parquet} \\
Bias-gap bootstrap samples & 1000 \\
Quality bootstrap samples & 20 \\
\bottomrule
\end{tabularx}
\caption{Milestone 3 experiment configuration.}
\end{table}

\noindent
Two findings are immediate. First, sampled decoding improves diversity
substantially relative to greedy. Greedy has almost no lexical diversity
(\texttt{distinct-2 = 0.002}) and an extreme repeated 3-gram rate
(\texttt{0.997}), while the highest-temperature setting reaches
\texttt{distinct-2 = 0.411} and lowers repeated 3-gram rate to \texttt{0.303}.
Second, the negative-regard gap does not disappear under sampling. For the
\texttt{description} prompt type, the Black man vs White woman gap remains
positive under every decoding family in this run.

\begin{table}[t]
\centering
\small
\begin{tabularx}{\linewidth}{@{}lccc@{}}
\toprule
Decoding config & Distinct-1 & Distinct-2 & Repeated 3-gram rate \\
\midrule
Greedy & 0.001 & 0.002 & 0.997 \\
Temperature 0.7 & 0.059 & 0.270 & 0.472 \\
Temperature 1.0 & 0.074 & 0.358 & 0.353 \\
Temperature 1.3 & 0.084 & 0.411 & 0.303 \\
Top-k 20 & 0.062 & 0.309 & 0.398 \\
Top-k 50 & 0.074 & 0.358 & 0.350 \\
Top-k 100 & 0.082 & 0.383 & 0.341 \\
Top-p 0.8 & 0.060 & 0.284 & 0.431 \\
Top-p 0.9 & 0.065 & 0.316 & 0.390 \\
Top-p 0.95 & 0.069 & 0.334 & 0.373 \\
\bottomrule
\end{tabularx}
\caption{Milestone 3 quality metrics by decoding configuration.}
\end{table}

\begin{table}[t]
\centering
\small
\begin{tabularx}{\linewidth}{@{}lcc@{}}
\toprule
Decoding config & Longest repetition span & 95\% CI \\
\midrule
Greedy & 0.0 & (0.0, 0.0) \\
Temperature 0.7 & 3.0 & (2.48, 3.0) \\
Temperature 1.0 & 4.0 & (3.0, 4.0) \\
Temperature 1.3 & 4.0 & (3.0, 4.0) \\
Top-k 20 & 2.0 & (2.0, 2.0) \\
Top-k 50 & 3.0 & (2.48, 3.0) \\
Top-k 100 & 4.0 & (3.0, 4.0) \\
Top-p 0.8 & 2.0 & (2.0, 2.0) \\
Top-p 0.9 & 2.0 & (2.0, 2.0) \\
Top-p 0.95 & 3.0 & (2.0, 3.0) \\
\bottomrule
\end{tabularx}
\caption{Milestone 3 longest-repetition-span summary.}
\end{table}

\begin{table}[t]
\centering
\small
\begin{tabularx}{\linewidth}{@{}lccc@{}}
\toprule
Decoding config & Description: Black man $-$ White woman $\Delta_{\text{neg}}$ & 95\% CI & Note \\
\midrule
Greedy & 0.250 & (0.198, 0.302) & Highest repetition, lowest diversity \\
Temperature 0.7 & 0.185 & (0.132, 0.238) & Gap remains positive \\
Temperature 1.0 & 0.193 & (0.142, 0.250) & Better diversity than temperature 0.7 \\
Temperature 1.3 & 0.133 & (0.078, 0.192) & Smallest gap in this trace \\
Top-k 20 & 0.205 & (0.152, 0.260) & Conservative top-k setting \\
Top-k 50 & 0.163 & (0.108, 0.217) & Middle top-k setting \\
Top-k 100 & 0.202 & (0.147, 0.262) & More diverse, still positive \\
Top-p 0.8 & 0.208 & (0.155, 0.262) & Strong gap remains \\
Top-p 0.9 & 0.152 & (0.098, 0.205) & Lower than top-p 0.8 \\
Top-p 0.95 & 0.185 & (0.128, 0.242) & Higher-entropy top-p run \\
\bottomrule
\end{tabularx}
\caption{Milestone 3 bias-gap trace for one consistent comparison. The sign stays positive under every decoding family in this run.}
\end{table}

\section{Personal Contributions}
\subsection*{Aleksandr Michailov}
Aleksandr Michailov completed the three reviewed repository commits that
established the Milestone 2 baseline:
\begin{itemize}
\item \texttt{818ce87}: report and documentation updates for the Milestone 2 handoff,
\item \texttt{06b9063}: refactor to the Pydantic-based application structure,
\item \texttt{2eec7eb}: scoring pipeline, metrics, sanity checks, and baseline reporting.
\end{itemize}
These commits cover the full Milestone 2 milestone, and the Milestone 2 deliverables are
complete and in good condition.

\subsection*{Ivan Chabanov}
Ivan Chabanov completed the Milestone 3 implementation and debugging work in the
current branch:
\begin{itemize}
\item decoding-grid generation and scoring commands,
\item quality metrics and bootstrap confidence intervals,
\item scorer stabilization and memory-safe loading,
\item cache compatibility, deduplication, and MPS-capable execution,
\item the combined Milestone 2--3 report and final result population.
\end{itemize}
These changes cover the full Milestone 3 milestone and produce the final decoding
sweep outputs used in this report.

\section{Artifact References}
The Milestone 3 results above are populated from the following generated artifacts:
\begin{itemize}
\item \texttt{outputs/metrics/a355f4e569239a813a35\_Milestone3\_combined\_Milestone3\_regard\_distributions.csv}
\item \texttt{outputs/metrics/a355f4e569239a813a35\_Milestone3\_combined\_Milestone3\_negative\_gaps\_with\_ci.csv}
\item \texttt{outputs/metrics/a355f4e569239a813a35\_Milestone3\_combined\_Milestone3\_quality\_metrics\_with\_ci.csv}
\item \texttt{outputs/metrics/a355f4e569239a813a35\_Milestone3\_combined\_Milestone3\_summary.json}
\end{itemize}

No large raw generations should be pasted into the report. If excerpts are
needed, include only minimal examples with an explicit content warning.

\end{document}
